TEMPERATURES AND METALLICITIES FOR 10,000+ M DWARFS IN THE \apogee\ SURVEY

### 2018-08-08 paper/draft.tex
commit all

\documentclass[preprint]{aastex62}

@@

@@

\title{APOGEE CATALOG OF 10,000+ M DWARFS MEASURED WITH DATA-DRIVEN MODELS}

@@

The Cannon \citep{Ness:2015} is a flexible, data-driven spectral modeling and parameter inference framework, demonstrated on high-resolution Apache Point Galactic Evolution Experiment (APOGEE; $\lambda/\Delta\lambda\sim22,500$, $1.5-1.7 \mu m$) spectra of giant stars to estimate stellar labels (T$_{\rm eff}$, logg, [Fe/H], and chemical abundances) to precisions higher than the model-grid pipeline. The lack of reliable stellar parameters reported by the APOGEE pipeline for temperatures less than $\sim3550K$ \citep{Schmidt:2016}, motivates the extension of this approach to M dwarf stars. Using a training sample of 87 M dwarfs ranging $2859 < T_{\rm eff} < 4131K$, and -0.48 $<$ [Fe/H] $<$ 0.49 we train a two-parameter model precise to 68K and 0.07 dex respectively. We then train a one-dimensional spectral type model using 51 M dwarfs ranging M0-M9 obtained from SDSS optical spectra, and demonstrate that The Cannon can infer spectral types to a precision of 0.6 types. Finally we apply our models to determine spectral subtypes, temperatures and metallicities for 10,084 M dwarfs in the survey and compare our results to APOGEE pipeline measurements, and find our results to be significantly more precise and accurate than ASPCAP, with scatters of $\sim80-100K$ comapred to color-temperature relations from several literature sources.

@@

The launch of the (APOGEE; \citealt{Majewski:2015}) survey as part of the Sloan Digital Sky Survey III mission \citealt{Eisenstein:2011} has introduced the largest sample of M dwarfs observed with high resolution spectroscopy. \color{gcolor}{What about APOGEE's wavelength coverage, resolution, selection and other specs allow us to do things with M dwarfs that other surveys/instruments can't?}\color{black}

A number of M dwarf studies have already been conducted in APOGEE (\citealt{Schmidt:2016} -- analyzed trends and reliablility of ASPCAP parameters for K/M dwarfs; \citealt{Souto:2017} -- modeled 2 M dwarfs, determined Teff/logg/metallicity + 13 elemental abundances; \citealt{Desphande:2013} -- modeled rotational and radial velocites for several hundred M dwarfs; \citealt{Rajpurohit:2018} -- tested BTSettl model grids on 45 M dwarfs, estimated Teff/logg/metallicity). \citealt{Gilhool:2018}; \citealt{Skinner:2018}.

Why is there a problem labeling M dwarfs (and especially for high res APOGEE spectra)? Emphasize issues: (1) models dont accurately reproduce data; (2) empirical standards are relative not absolute, and they are sparesely sampled. Several approaches to to determining physical parameters and the limitations:

@@

Why The Cannon approach? brief importance statment cannon and what it does (cite \citealt{Ness:2015}, \citealt{Ho:2017a}, \citealt{Casey:2016}) \color{gcolor}{For Hogg to articulate...}\color{black}

@@

\item Make the point that The Cannon does not require us to have good M-dwarf models in the APOGEE wavelength range at all

@@

\subsection{APOGEE Survey}
% \begin{enumerate}
% Describe the APOGEE data (survey mission: \citealt{Majewski:2015}; target selection: \citealt{Zasowski:2013}; ASPCAP pipeline: \citealt{Perez:2016}).

% \item[-] Some M dwarfs in APOGEE proposed as ancilliary targets (Blake/Covey sources), plus an unknown number observed unintentionally \citep{Desphande:2013}.
% \end{enumerate}

The APOGEE survey \citep{Majewski:2015} of the Sloan Digital Sky Survey III and IV (SDSS-III/IV \citealt{Eisenstein:2011}; \citealt{Blanton:2017}) is a high resolution (R$\sim$22,500), H-band (1.5-1.7$\mu$m) survey which has observed over 260,000 stellar spectra since its fourteenth data release (DR14; \citealt{Abolfathi:2017}). Fundamental parameters for each of these stars are estimated by the APOGEE Stellar Parameter and Chemical Abundances Pipeline (ASPCAP; \citealt{Perez:2016}), which employs a $\chi^2$ fitting procedure using the FERRE code to fit ATLAS9 radiative transfer models \citep{Castelli:2004} to determine atmospheric parameters, 15 chemical abundances and micro-turbulence parameters (T$_{\rm eff}$, log g, [Fe/H], [$\alpha$/M], [C/M], [N/M]; \citealt{Meszaros:2012}).

APOGEE is primarily targeted for bright stellar populations, with de-reddened photometry and color cutoffs of $7 \leq H \leq 13.8$ and $[J-K]_0 \geq 0.5$ \citep{Zasowski:2013} with the objective of studying galactic composition and evolution. However numerous cool, main sequence sources have also been observed either as targets proposed by the M dwarf ancilliary survey (1,350 sources selected by proper motions from the  \citealt{Desphande:2013}), or randomly by APOGEE, which we quantify in this study to be over 10,000 in total.

@@

The Cannon is a \emph{generative model} which parameterizes the flux at each pixel of a spectrum in terms of a set of stellar labels (a flexible number of parameters chosen by the user; described in more detail in Section \ref{sec:cannon}). The model in this sense is used to \emph{transfer} labels from spectra which we know parameters for to those which we do not. Hence this data-driven approach effectively removes the challenges of physically modeling the atmosphere of a star (and common issues associated such as incomplete line lists or opacities), provided that we have a subset of spectra in the dataset with known (ideally very accurately measured) \emph{reference labels} that can be measured from other surveys/methods that are easier or more accurate to infer from.

@@

APOGEE data products: {\tt\string ap1D}, {\tt\string apVisit}, {\tt\string apStar}, and {\tt\string aspcapStar}. We use the {\tt\string aspcapStar} spectra, which contains the continuum-normalized, rest frame shifted, co-added spectrum of all observed epochs (see \citealt{Perez:2016} for a complete description of the pipeline). SNR of stars in our sample...

@@

\subsection{The Cannon: A Data-Driven Approach}

The Cannon is a fully empirical model, which applies no use of line lists or radiative transfer models. Describe the assumptions of the model: (1) Sources with identical labels have near-identical flux at each pixel; (2) Expected flux at each pixel varies continuosly with change in label.

The Cannon model can in principle be trained on any physical or empirical labels available. However, limitations to consider are that test labels from The Cannon are only accurate if the training labels accurate, and only precise if the training labels are measured using a consistent method. The Cannon does not extrapolate well outside the parameter space of the training sample.

@@

Another mode of analysis we can look at with The Cannon is how the derivative of the model changes with respect to given training parameter, which makes our model interpretable for discovering or verifying atomic or molecular lines with strong dependence on different physical parameters. As discussed further in sections \ref{subsec:mann_results} and \ref{subsec:west_results}, we show the derivative plots of each of our two models and compute the error of the derivative at each pixel using a jackknife statistic:
\begin{equation}
	\sigma_{\theta,m}^2 = \frac{N-1}{N} \sum^N_{n=1} (\theta_{/n}^m - \theta^m)^2 
\end{equation}
where $\sigma_{\theta,m}$ is the error at pixel \emph{m}, \emph{N} is the total number of stars in the sample, indexed by \emph{n}, $\theta$ is the coeffient vector trainined on all N sources, and $\theta_{/n}$ is the coefficient vector trained on \emph{N-1} sources excluding star \emph{n}.

\subsubsection{Temperature/Metallicity Model \label{subsec:mann_results}}
For the physical parameter model we trained The Cannon on 87 M dwarfs with a two-dimensional temperature/metallicity labels, to a precision of 68K/0.07dex respectively, similar to the original training sample uncertainties of 60K/0.08dex. We note for this model 6 out of 87 sources show strong visible rotational line broadening (as indicated by the red circles in Figure \ref{fig:mann_validation}), while the remaining sources are mostly old ($>8$ Gyr) and likely slowly rotating sources. While these sources with high rotational broadening have high $\chi^2$ values (those sources with $\chi^2 > 80,000$ in Figure \ref{fig:chi_dist}), we find that the scatter of labels derived for these sources is consistent with the rest of the sample and we do not remove them from the training sample.

@@

\subsubsection{Spectral Type Model \label{subsec:west_results}}
We trained The Cannon on 51 M dwarfs with a one-dimensional spectral type label, and obtained a precision of $\pm$0.6 spectral types, more precise than the original training label uncertainty of $\pm$1 spectral type. We note however, that the training sample is distributed heavily towards earlier type sources, with a median spectral type of 3 and only one M8 and one M9 source. As seen in Figure \ref{fig:west_validation}, the model performs well at re-deriving the labels for the M8 and M9 source when they are included in the model (less than 1 SPT difference), but badly when left out (over 2 SPT difference), which indicates The Cannon does a poor job at extrapolating to labels outside of the training sample space.

@@

\subsection{Test Sample \label{subsec:test_selection}} 
Out of the total APOGEE DR14 catalog of 258,475 sources, we selected 254,478 sources which were in the cross match of Gaia DR2 (\citealt{Brown:2018}) and applied Gaia color-magnitude cuts of $1<BP-RP<6$ and $7.5<M_G<20$, yielding a sample of 14,827 sources. From there we applied the following selection procedure to identify a sample of single, main-sequence M stars. Although our catalog reports the temperatures, metallicities, and spectral types for XXX sources out of 14,000 with $\chi^2<200,000$, the sources we deem ``safe'', with minimal contamination from reddened K dwarfs, pre-main sequence stars, and binaries, are selected by the following criteria (see Figure \ref{fig:cmd_selection}):

@@

\item \textbf{K dwarf cut:} In 2MASS photometry we select $M_K > 4.3$ to remove likely K dwarfs, which appear reddened in Gaia $BP$, $RP$ photometry.

\item \textbf{Pre-main sequence, multiple system and subdwarf (?) cut:} in Gaia and 2MASS we apply $.7<J-K<1.1$ and points in the selction region within the points $\left\{ (1.4, 7.5), (2.2, 7.5), (4.2, 14), (3.3, 14) \right\}$ of ($BP-RP$, $M_G$) to cut remaining likely pre-main sequence, multiple sources and subdwarfs.

@@

\item \textbf{Training bound selection:} Because extrapolation with The Cannon is unreliable, we select only sources with labels inside the range of the training sample: for the reported temperatures and metallicities we impose $2850<T_{\rm eff}<4150$ and $-0.5<[Fe/H]<0.5$, and for reported spectral subtypes we impose $0<SPT<9$.
\end{enumerate}

@@

In \citealt{Schmidt:2016} the authors used three different color-temperature relations to check the consistency of ASPCAP temperature values, using a combination of 2MASS and visual band photometries. \color{gcolor}{Brief description of the color-teff relations from \citealt{Casagrande:2008}, \citealt{Boyajian:2013}, and \citealt{Mann:2015}...}\color{black}.

To obtain visual band magnitudes for a set of sources, we cross-matched the 10,084 sources in our ``safe'' test sample to the AAVSO Photometric All-Sky Survey DR9 (APASS; \citealt{Henden:2016}), to obtain a subsample of 6,000 sources with both $BV$ photometries measured by APASS, and $JHK$ photometries from APOGEE. In Figure \ref{fig:aspcap_cannon_validation} we repeat the temperature validation tests for each of the three relations and compare the ASPCAP temperatures for the same sources. \color{gcolor}{Comments about validation results and the uncertainty of our Cannon temperatures...}\color{black}

@@

\begin{enumerate}
\item[-] Brief summary of what we did and found{}

@@

\item[-] \color{gcolor}{HOGG: Some discussion of precision and accuracy: The Cannon can produce precise results, but is only as accurate as its training set.}\color{black}

@@

\item[-] The Cannon as a test of evaluating training parameter accuracy? \color{gcolor}{HOGG: Should this be precision, not accuracy? For example adding a constant factor (say adding +1000 to every metallicity training value) does not change the scatter of the test labels}\color{black}

@@

%---------------------------
\begin{figure}[ht]
\begin{center}
\includegraphics[width=16cm]{figures/demo_derivatives_teff1.png}
\includegraphics[width=16cm]{figures/demo_derivatives_teff2.png}
\includegraphics[width=16cm]{figures/demo_derivatives_teff3.png}
\end{center}
\caption{\textit{Top two panels of each plot:} Mann-trained model for varying temperatures, and similar metallcities. \textit{Third panel of each plot:} Derivative of the Cannon model with respect to temperature, taken at the median training temperature, T$_{\rm eff}=3463 K$}. \label{fig:demo_teff}
\end{figure}

%---------------------------
\begin{figure}[ht]
\begin{center}
\includegraphics[width=16cm]{figures/demo_derivatives_feh1.png}
\includegraphics[width=16cm]{figures/demo_derivatives_feh2.png}
\includegraphics[width=16cm]{figures/demo_derivatives_feh3.png}
\end{center}
\caption{\textit{Top two panels of each plot:} Mann-trained model for varying metallicities and similar temperatures. \textit{Third panel:} Derivative of the Cannon model with respect to metallicity, taken at the median training metallicity, [Fe/H]$=-0.03$.} \label{fig:demo_feh}
\end{figure}

%---------------------------
\begin{figure}[ht]
\begin{center}
\includegraphics[width=12cm]{figures/Spectral_Sequence_1.png}
\includegraphics[width=12cm]{figures/Spectral_Sequence_2.png}
\includegraphics[width=12cm]{figures/Spectral_Sequence_3.png}
\end{center}
\caption{Spectral sequence of dwarfs in training set M0-M9; separate plots show three detector chips of APOGEE spectrum (black), the best fit Cannon trained model (red) with highlighted spectral type sensitive regions identified in \citealt{Desphande:2013}.} \label{fig:sp_sequence}
\end{figure}

@@

\begin{figure}[ht]
\plotone{figures/cannon_chi_dist.png}
\caption{\textit{Left:} the distribution of fits for the training sample and known samples of pre-MS stars (\citealt{Cottaar:2014}) and binary sources (\citealt{ElBadry:2018}; \citealt{Skinner:2018}); \textit{right:} the distribution of $\chi^2$ fits for all 14,827 sources in the APOGEE-Gaia cross match, with color cuts $1<G_{BP}-G_{RP}<6$ and $7.5<M_{G}<20$ and $\pi>0$. We apply a quality cut of $\chi^2 < 100,000$ to the test sample for those sources we report as "safe". \label{fig:chi_dist}}
\end{figure}

\begin{figure}[ht]
\plotone{figures/cmd_selection.png}
\caption{Gaia and 2MASS color-magnitude cuts for 12,037 sources with $\chi^2<100,000$. Over plotted with orange triangles are the 67 out of 87 training sample sources which have parallaxes measured by Gaia. \label{fig:cmd_selection}}
\end{figure}

\begin{figure}[ht]
\plotone{figures/safe_selection.png}
\caption{Final sample selection with 10,084 M dwarfs based on criteria described in Section \ref{subsec:test_selection}. \label{fig:safe_selection}}
\end{figure}

\begin{figure}[ht]
\plotone{figures/aspcap_cannon_label_hist.png}
\caption{Distribution of temperatures and metallicities for 10,084 sources from The Cannon compared to the ASPCAP pipeline. \label{fig:aspcap_cannon_label_hist}}
\end{figure}

@@

% \startlongtable{}
% \begin{deluxetable*}{ccc|ccc|ccc}
% \tablecaption{Cannon results for Temperature/Metallicity model \label{mann_results}}
% \tabletypesize{\scriptsize}
% \tablehead{
% \multicolumn{3}{c}{\underline{Designation}} & \multicolumn{3}{c}{\underline{Temperature (K)}} & \multicolumn{3}{c}{\underline{Metallicity (dex)}} \\
% 	  \colhead{2MASS ID} & \colhead{RA}   & \colhead{DEC}
%     & \colhead{Training} & \colhead{Test} & \colhead{LOOCV}
%     & \colhead{Training} & \colhead{Test} & \colhead{LOOCV}  
% } 
% \startdata
% 2M00182256+4401222 & 4.59542   & 44.02278  & 3603        & 3532       & 3519        & -0.3       & -0.27     & -0.24      \\
% 2M00182549+4401376 & 4.60779   & 44.02734  & 3218        & 3527       & 3554        & -0.3       & -0.28     & -0.27      \\
% 2M00285391+5022330 & 7.22488   & 50.37588  & 3207        & 3185       & 3188        & 0.11       & 0.05      & 0.02       \\
% 2M00401001+0308050 & 10.04169  & 3.13473   & 3725        & 3783       & 3778        & 0.04       & 0.11      & 0.13       \\
% 2M00580115+3919111 & 14.50482  & 39.31977  & 3157        & 3094       & 3101        & -0.07      & -0.01     & 0.03       \\
% 2M01232542+1638384 & 20.8559   & 16.64401  & 3272        & 3218       & 3217        & 0.1        & 0.03      & 0.0        \\
% 2M02001278+1303112 & 30.05402  & 13.05196  & 3080        & 3057       & 3074        & -0.16      & -0.1      & -0.01      \\
% 2M02361535+0652191 & 39.06358  & 6.87167   & 3284        & 3249       & 3253        & -0.12      & -0.29     & -0.29      \\
% 2M03044335+6144097 & 46.18104  & 61.73583  & 3500        & 3468       & 3467        & -0.12      & -0.26     & -0.27      \\
% 2M03553688+5214291 & 58.90373  & 52.2414   & 3435        & 3381       & 3369        & -0.35      & -0.25     & -0.18      \\
% 2M04125880+5236421 & 63.24499  & 52.61165  & 3100        & 3181       & 3200        & -0.04      & -0.01     & 0.01       \\
% 2M04310001+3647548 & 67.75003  & 36.79855  & 3419        & 3371       & 3369        & 0.08       & 0.11      & 0.1        \\

>>>>

\documentclass[modern]{aastex62}

@@

\newcommand{\apogee}{\textsl{APOGEE}}
\newcommand{\thecannon}{\textsl{The Cannon}}
\newcommand{\cannon}{\textsl{Cannon}}
\newcommand{\aspcap}{\textsl{ASPCAP}}
\newcommand{\gaia}{\textsl{Gaia}}
\newcommand{\zmass}{\textsl{2MASS}}
\newcommand{\apass}{\textsl{APASS}}
\newcommand{\sdssiv}{\textsl{SDSS-IV}}

@@

\title{TEMPERATURES AND METALLICITIES FOR 10,000+ M DWARFS IN THE \apogee\  SURVEY}

@@

\thecannon\ \citep{Ness:2015} is a flexible, data-driven spectral modeling and parameter inference framework, demonstrated on high-resolution Apache Point Galactic Evolution Experiment (\apogee; $\lambda/\Delta\lambda\sim22,500$, $1.5-1.7 \mathrm{\mu m}$) spectra of giant stars to estimate stellar labels (T$_{\rm eff}$, logg, [Fe/H]) to precisions higher than the model-grid pipeline. The lack of reliable stellar parameters reported by the \apogee\ pipeline for temperatures less than $\sim3550$K \citep{Schmidt:2016}, motivates the extension of this approach to M dwarf stars. Using a training sample of 87 M dwarfs ranging $2859 < T_{\rm eff} < 4131$K, and $-0.48 <$ [Fe/H] $< 0.49$ dex we train a two-parameter model precise to 68K and 0.07 dex respectively. We then train a one-dimensional spectral type model using 51 M dwarfs ranging M0-M9 obtained from SDSS optical spectra, and demonstrate that \thecannon\ can infer spectral types to a precision of 0.6 types. Finally we apply our models to determine spectral subtypes, temperatures and metallicities for 10,311 M dwarfs in the survey, and compare our results to \apogee\ pipeline measurements, and find our results to be significantly more precise and less biased than \aspcap\, with scatters of $\sim80-100$K comapred to color-temperature relations from several literature sources.

@@

The launch of the (\apogee\; \citealt{Majewski:2015}) survey as part of the Sloan Digital Sky Survey III mission \citealt{Eisenstein:2011} has introduced the largest sample of M dwarfs observed with high resolution spectroscopy. \color{gcolor}{What about \apogee\'s wavelength coverage, resolution, selection and other specs allow us to do things with M dwarfs that other surveys/instruments can't?}\color{black}

A number of M dwarf studies have already been conducted in \apogee\ (\citealt{Schmidt:2016} -- analyzed trends and reliablility of ASPCAP parameters for K/M dwarfs; \citealt{Souto:2017} -- modeled 2 M dwarfs, determined Teff/logg/metallicity + 13 elemental abundances; \citealt{Desphande:2013} -- modeled rotational and radial velocites for several hundred M dwarfs; \citealt{Rajpurohit:2018} -- tested BTSettl model grids on 45 M dwarfs, estimated Teff/logg/metallicity). \citealt{Gilhool:2018}; \citealt{Skinner:2018}.

Why is there a problem labeling M dwarfs (and especially for high res \apogee\ spectra)? Emphasize issues: (1) models dont accurately reproduce data; (2) empirical standards are relative not absolute, and they are sparesely sampled. Several approaches to to determining physical parameters and the limitations:

@@

Why \thecannon\ approach? brief importance statment cannon and what it does (cite \citealt{Ness:2015}, \citealt{Ho:2017a}, \citealt{Casey:2016}) \color{gcolor}{For Hogg to articulate...}\color{black}

@@

\item Make the point that \thecannon\ does not require us to have good M-dwarf models in the \apogee\ wavelength range at all

@@

\subsection{\apogee\ Survey}
% \begin{enumerate}
% Describe the \apogee\ data (survey mission: \citealt{Majewski:2015}; target selection: \citealt{Zasowski:2013}; ASPCAP pipeline: \citealt{Perez:2016}).

% \item[-] Some M dwarfs in \apogee\ proposed as ancilliary targets (Blake/Covey sources), plus an unknown number observed unintentionally \citep{Desphande:2013}.
% \end{enumerate}

The \apogee\ survey \citep{Majewski:2015} of the Sloan Digital Sky Survey III and IV (SDSS-III/IV \citealt{Eisenstein:2011}; \citealt{Blanton:2017}) is a high resolution (R$\sim$22,500), H-band (1.5-1.7$\mu$m) survey which has observed over 260,000 stellar spectra since its fourteenth data release (DR14; \citealt{Abolfathi:2017}). Fundamental parameters for each of these stars are estimated by the \apogee\ Stellar Parameter and Chemical Abundances Pipeline (ASPCAP; \citealt{Perez:2016}), which employs a $\chi^2$ fitting procedure using the FERRE code to fit ATLAS9 radiative transfer models \citep{Castelli:2004} to determine atmospheric parameters, 15 chemical abundances and micro-turbulence parameters (T$_{\rm eff}$, log g, [Fe/H], [$\alpha$/M], [C/M], [N/M]; \citealt{Meszaros:2012}).

\apogee\ is primarily targeted for bright stellar populations, with de-reddened photometry and color cutoffs of $7 \leq H \leq 13.8$ and $[J-K]_0 \geq 0.5$ \citep{Zasowski:2013} with the objective of studying galactic composition and evolution. However numerous cool, main sequence sources have also been observed either as targets proposed by the M dwarf ancilliary survey (1,350 sources selected by proper motions from the  \citealt{Desphande:2013}), or randomly by \apogee\, which we quantify in this study to be over 10,000 in total.

@@

\thecannon\ is a \emph{generative model} which parameterizes the flux at each pixel of a spectrum in terms of a set of stellar labels (a flexible number of parameters chosen by the user; described in more detail in Section \ref{sec:cannon}). The model in this sense is used to \emph{transfer} labels from spectra which we know parameters for to those which we do not. Hence this data-driven approach effectively removes the challenges of physically modeling the atmosphere of a star (and common issues associated such as incomplete line lists or opacities), provided that we have a subset of spectra in the dataset with known (ideally very accurately measured) \emph{reference labels} that can be measured from other surveys/methods that are easier or more accurate to infer from.

@@

\apogee\ data products: {\tt\string ap1D}, {\tt\string apVisit}, {\tt\string apStar}, and {\tt\string aspcapStar}. We use the {\tt\string aspcapStar} spectra, which contains the continuum-normalized, rest frame shifted, co-added spectrum of all observed epochs (see \citealt{Perez:2016} for a complete description of the pipeline). SNR of stars in our sample...

@@

\subsection{\thecannon\: A Data-Driven Approach}

\thecannon\ is a fully empirical model, which applies no use of line lists or radiative transfer models. Describe the assumptions of the model: (1) Sources with identical labels have near-identical flux at each pixel; (2) Expected flux at each pixel varies continuosly with change in label.

\thecannon\ model can in principle be trained on any physical or empirical labels available. However, limitations to consider are that test labels from \thecannon\ are only accurate if the training labels accurate, and only precise if the training labels are measured using a consistent method. \thecannon\ does not extrapolate well outside the parameter space of the training sample.

@@

Another mode of analysis we can look at with \thecannon\ is how the derivative of the model changes with respect to given training parameter, which makes our model interpretable for discovering or verifying atomic or molecular lines with strong dependence on different physical parameters. As discussed further in sections \ref{subsec:mann_results} and \ref{subsec:west_results}, we show the derivative plots of each of our two models and compute the error of the derivative at each pixel using a jackknife statistic:
\begin{equation}
	\sigma_{\theta,m}^2 = \frac{N-1}{N} \sum^N_{n=1} (\theta_{/n}^m - \theta^m)^2 
\end{equation}
where $\sigma_{\theta,m}$ is the error at pixel \emph{m}, \emph{N} is the total number of stars in the sample, indexed by \emph{n}, $\theta$ is the coeffient vector trainined on all N sources, and $\theta_{/n}$ is the coefficient vector trained on \emph{N-1} sources excluding star \emph{n}.

\subsubsection{Temperature/Metallicity Model \label{subsec:mann_results}}
For the physical parameter model we trained \thecannon\ on 87 M dwarfs with a two-dimensional temperature/metallicity labels, to a precision of 77K/0.09dex as estimated by the cross validation scatter, similar to the original training sample uncertainties of 60K/0.08dex. We note for this model 6 out of 87 sources show strong visible rotational line broadening (as indicated by the red circles in Figure \ref{fig:mann_validation}), while the remaining sources are mostly old ($>8$ Gyr) and likely slowly rotating sources. While these sources with high rotational broadening have high $\chi^2$ values (those sources with $\chi^2 > 80,000$ in Figure \ref{fig:chi_dist}), we find that the scatter of labels derived for these sources is consistent with the rest of the sample and we do not remove them from the training sample.

@@

\subsubsection{Spectral Type Model \label{subsec:west_results}}
We trained \thecannon\ on 51 M dwarfs with a one-dimensional spectral type label, and obtained a precision of $\pm$0.6 spectral types, more precise than the original training label uncertainty of $\pm$1 spectral type. We note however, that the training sample is distributed heavily towards earlier type sources, with a median spectral type of 3 and only one M8 and one M9 source. As seen in Figure \ref{fig:west_validation}, the model performs well at re-deriving the labels for the M8 and M9 source when they are included in the model (less than 1 SPT difference), but badly when left out (over 2 SPT difference), which indicates \thecannon\ does a poor job at extrapolating to labels outside of the training sample space.

@@

\subsection{Test Sample \label{subsec:test_selection}} 
Out of the total \apogee\ DR14 catalog of 258,475 sources, we selected 254,478 sources which were in the cross match of \textsl{Gaia DR2} (\citealt{Brown:2018}) and applied \gaia\ color-magnitude cuts of $1<BP-RP<6$ and $7.5<M_G<20$, yielding a sample of 14,827 sources. From there we applied the following selection procedure to identify a sample of ``safe'' single, main-sequence M stars, with minimal contamination from reddened K dwarfs, pre-main sequence stars, and binaries, are selected by the following criteria (see Figure \ref{fig:cmd_selection}):

@@

\item \textbf{K dwarf cut:} In \zmass\ photometry we select $M_K > 4.3$ to remove likely K dwarfs, which appear reddened in \gaia\ $BP$, $RP$ photometry.

\item \textbf{Pre-main sequence, multiple system and subdwarf (?) cut:} in \gaia\ and \zmass\ we apply $.7<J-K<1.1$ and points in the selction region within the points $\left\{ (1.4, 7.5), (2.2, 7.5), (4.2, 14), (3.3, 14) \right\}$ of ($BP-RP$, $M_G$) to cut remaining likely pre-main sequence, multiple sources and subdwarfs.

@@

\item \textbf{Training bound selection:} Because extrapolation with \thecannon\ is unreliable, we select only sources with labels inside the range of the training sample: for the reported temperatures and metallicities we impose $2850<T_{\rm eff}<4150$ and $-0.5<[Fe/H]<0.5$, and for reported spectral subtypes we impose $0<SPT<9$.
\end{enumerate}

@@

In \citealt{Schmidt:2016} the authors used three different color-temperature relations to check the consistency of \aspcap\ temperature values using a combination of \zmass\ and visual band photometries. \color{gcolor}{Brief description of the color-teff relations from \citealt{Casagrande:2008}, \citealt{Boyajian:2013}, and \citealt{Mann:2015}...}\color{black}. To obtain visual band magnitudes for a set of sources, we cross-matched the 10,311 sources in our ``safe'' test sample to the AAVSO Photometric All-Sky Survey DR9 (\apass; \citealt{Henden:2016}), to obtain a subsample of 6,000 sources with both $BV$ photometries measured by \apass, and $JHK$ photometries from \apogee. In Figure \ref{fig:aspcap_cannon_validation} we repeat the temperature validation tests for each of the three relations and compare the \aspcap\ temperatures for the same sources. \color{gcolor}{Comments about validation results and the uncertainty of our Cannon temperatures...}\color{black}

The gradients in the color magnitude diagrams of Figure \ref{fig:safe_selection} demonstrate the expected negative correlation between empirical \gaia\ $BP-RP$ colors and West-trained spectral classifications with the physical Mann-trained temperatures. Spectral type decreases nonlinearly with temperature... negatively correlated with metallcity for spectral types hotter than M3

@@

\begin{enumerate}
\item[-] Brief summary of what we did and found...

@@

\item[-] \color{gcolor}{HOGG: Some discussion of precision and accuracy: \thecannon\ can produce precise results, but is only as accurate as its training set.}\color{black}

@@

\item[-] \thecannon\ as a test of evaluating training parameter accuracy?

@@

%---------------------------
\begin{figure}[ht]
\begin{center}
\includegraphics[width=16cm]{figures/demo_derivatives_teff1.png}
\includegraphics[width=16cm]{figures/demo_derivatives_teff2.png}
\includegraphics[width=16cm]{figures/demo_derivatives_teff3.png}
\end{center}
\caption{\textit{Top two panels of each plot:} Mann-trained model for varying temperatures, and similar metallcities. \textit{Third panel of each plot:} Derivative of \thecannon\ model with respect to temperature, taken at the median training temperature, T$_{\rm eff}=3463 K$}. \label{fig:demo_teff}
\end{figure}

%---------------------------
\begin{figure}[ht]
\begin{center}
\includegraphics[width=16cm]{figures/demo_derivatives_feh1.png}
\includegraphics[width=16cm]{figures/demo_derivatives_feh2.png}
\includegraphics[width=16cm]{figures/demo_derivatives_feh3.png}
\end{center}
\caption{\textit{Top two panels of each plot:} Mann-trained model for varying metallicities and similar temperatures. \textit{Third panel:} Derivative of \thecannon\ model with respect to metallicity, taken at the median training metallicity, [Fe/H]$=-0.03$.} \label{fig:demo_feh}
\end{figure}

%---------------------------
\begin{figure}[ht]
\begin{center}
\includegraphics[width=12cm]{figures/Spectral_Sequence_1.png}
\includegraphics[width=12cm]{figures/Spectral_Sequence_2.png}
\includegraphics[width=12cm]{figures/Spectral_Sequence_3.png}
\end{center}
\caption{Spectral sequence of dwarfs in training set M0-M9; separate plots show three detector chips of \apogee\ spectrum (black), the best fit Cannon trained model (red) with highlighted spectral type sensitive regions identified in \citealt{Desphande:2013}.} \label{fig:sp_sequence}
\end{figure}

@@

\begin{figure}[ht]
\plotone{figures/cannon_chi_dist.png}
\caption{\textit{Left:} the distribution of fits for the training sample and known samples of pre-MS stars (\citealt{Cottaar:2014}) and binary sources (\citealt{ElBadry:2018}; \citealt{Skinner:2018}); \textit{right:} the distribution of $\chi^2$ fits for all 14,827 sources in the \apogee\-\gaia\ cross match, with color cuts $1<G_{BP}-G_{RP}<6$ and $7.5<M_{G}<20$ and $\pi>0$. We apply a quality cut of $\chi^2 < 100,000$ to the test sample for those sources we report as "safe". \label{fig:chi_dist}}
\end{figure}

\begin{figure}[ht]
\plotone{figures/cmd_selection.png}
\caption{\gaia\ and \zmass\ color-magnitude cuts for 12,037 sources with $\chi^2<100,000$. Over plotted with orange triangles are the 67 out of 87 training sample sources which have parallaxes measured by \gaia. \label{fig:cmd_selection}}
\end{figure}

\begin{figure}[ht]
\gridline{\fig{figures/safe_selection_teff.png}{0.5\textwidth}{}
          \fig{figures/safe_selection_feh.png}{0.5\textwidth}{}}
\gridline{\fig{figures/safe_selection_spt.png}{0.5\textwidth}{}}
\caption{Final sample selection with 10,311 M dwarfs based on criteria described in Section \ref{subsec:test_selection}. \label{fig:safe_selection}}
\end{figure}

\begin{figure}[ht]
\plotone{figures/spt_vs_teff_vs_feh.png}
\caption{West-trained spectral types vs. Mann-trained metallcities and temperatures. For each integer spectral type bin the mean temperatures and metallicities are shown as red error boxes. \label{fig:west_vs_mann_labels}}
\end{figure}

\begin{figure}[ht]
\plotone{figures/aspcap_cannon_label_hist.png}
\caption{Distribution of temperatures and metallicities for 10,311 sources from \thecannon\ compared to the \aspcap\ pipeline. \label{fig:aspcap_cannon_label_hist}}
\end{figure}

@@

\begin{longrotatetable}
\begin{deluxetable*}{ccc|ccc|ccc|c}
\tablecaption{Cannon results for Temperature/Metallicity model; see online journal for full table. \label{mann_results}}
\tabletypesize{\scriptsize}
\tablehead{
\multicolumn{3}{c}{\underline{Designation}} & \multicolumn{3}{c}{\underline{Temperature (K)}} & \multicolumn{3}{c}{\underline{Metallicity (dex)}} & \multicolumn{1}{c}{\underline{Model Fit}} \\
	  \colhead{2MASS ID} & \colhead{RA}   & \colhead{DEC}
    & \colhead{Training} & \colhead{Test} & \colhead{LOOCV}
    & \colhead{Training} & \colhead{Test} & \colhead{LOOCV} 
    & \colhead{Test $\chi^2$} 
} 
\startdata
2M00182256+4401222 & 4.59542   & 44.02278  & 3603        & 3538       & 3525        & -0.3       & -0.28     & -0.26      & 15304   \\
2M00182549+4401376 & 4.60779   & 44.02734  & 3218        & 3528       & 3560        & -0.3       & -0.29     & -0.29      & 21273   \\
2M00285391+5022330 & 7.22488   & 50.37588  & 3207        & 3190       & 3192        & 0.11       & 0.04      & 0.01       & 31790   \\
2M00401001+0308050 & 10.04169  & 3.13473   & 3725        & 3777       & 3772        & 0.04       & 0.12      & 0.14       & 9830    \\
2M00580115+3919111 & 14.50482  & 39.31977  & 3157        & 3100       & 3107        & -0.07      & -0.02     & 0.01       & 13337   \\
2M01232542+1638384 & 20.8559   & 16.64401  & 3272        & 3225       & 3223        & 0.1        & 0.02      & -0.01      & 34797   \\
2M02001278+1303112 & 30.05402  & 13.05196  & 3080        & 3059       & 3077        & -0.16      & -0.1      & -0.02      & 33210   \\
2M02361535+0652191 & 39.06358  & 6.87167   & 3284        & 3241       & 3243        & -0.12      & -0.28     & -0.28      & 48445   \\
2M03044335+6144097 & 46.18104  & 61.73583  & 3500        & 3466       & 3466        & -0.12      & -0.26     & -0.28      & 40739   \\
2M03553688+5214291 & 58.90373  & 52.2414   & 3435        & 3386       & 3375        & -0.35      & -0.28     & -0.2       & 53870   \\
2M04125880+5236421 & 63.24499  & 52.61165  & 3100        & 3183       & 3204        & -0.04      & -0.01     & 0.0        & 46007   \\
